---
title: "main"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
```

Narative: 

What musical characteristics of  an electronic song make it popular? This is the question I aim to answer in my project. 

By characteristics of a song, I mean the basic features in the context of music theory such as tempo, key, time signature, etc. But I also mean more complex features in the context of how humans interpret music such as, loudness, energy, danceability, etc.

By popularity of a song, I mean how many times that song was listened to compared to other songs. 

For music producers, an accurate answer to this question would be very valuable as it would give them evidence for selecting target characteristics to maximize a songs popularity. It would also give them target characteristics for a certain type of song. For example, if a music producer wanted to create a high energy and very danceable song, they would have evidence as to which key, time signature, tempo, etc. they would need to maximize the songs popularity. 

My response variable would be the popularity of a song based on its musical characteristics. However, my dataset only contains time-series measures of popularity. That is, my dataset only contains information of how popular a song is currently but I am more interested in how popular a song is independent of when it was uploaded. The songs popularity in the dataset is also dependent on the popularity of the artists from the assumption that more popular artists have bigger platforms to promote their song. I would like my response variable to also be independent of the popularity of the artists. Therefor I need to construct a custom response variable.

To tackle the issue of the current popularity metric from the data-set being a time dependent variable. The response variable would be the time dependent popularity metric but it's predicted value would be calculated based on a given time and artist popularity.  

This article :https://towardsdatascience.com/song-popularity-predictor-1ef69735e380 did a very similar thing to my project. Only it used the song popularity as a predictor variable to estimate the song features. That is, it analyzed the distribution of popular songs to predict the features of those songs. 


My hypothesis is that a songs popularity is dependedent on its musical characteristics. 



Data:

This dataset is retreived from Spotify's API. See the script below. 

Script: https://colab.research.google.com/drive/1VIpipujOuut-qT1iVkMw8lASIqxjEGDv?usp=sharing


The dataset was created by selecting playlists within the Electronic Music Genre, then calling the API to retreive track information for every track within every playlist. Then for every track I called the API again to get information that was not included in the first call. This is information like artist information and upload date. I then parsed everything into a dictionary and wrote it to the CSV file. 

I may need to do some more data augmentation. For example, I may want to augment some of the predictor variables I imagine to have no linear relationship with the response based on the context of the problem to be categorical to account for this non-linear relationship. 

This article: https://towardsdatascience.com/song-popularity-predictor-1ef69735e380 used logistic regression which is evidence that categorical variables may be more accurate predictors. It also did a fair bit more cleaning of the dataset. Which is evidence that I may need to look at my data more closely for corrupted entries. 



```{r}
track_data <- read.csv('data/spotify_tracks.csv')
```
```{r}
glimpse(track_data)
summary(track_data)
```
```{r}
t_popularity <- ggplot(track_data, aes(x=track.popularity)) + geom_histogram()
t_popularity
```
The track popularity seems very normal, asside from the outliers at 0 popularity. Maybe I can change my problem with the given that the track popularity will not be 0? This would allow me to remove such outliers. 

Lets see if there may be an ascociation between track popularity and the average artist popularity. 
```{r}
popart <- ggplot(data = track_data, aes(x = pop/num.artist , y = track.popularity)) + geom_point()+ stat_smooth(method = "lm", col = "red")
popart
```

As expected, we can observe a possible association between the popularity of tracks and their artists. We may need to investigate the outlier tracks with popularity is 0. 



Lets plot the number of artists on a track. 
```{r}
t_nart <- ggplot(track_data, aes(x=num.artist)) + geom_bar()
t_nart
```
We can see that the vast majority of number of artists for a track are between [1,3]
Lets plot the sum popularity for tracks with number of artists between [1,3]





```{r}
data_nart1 <- track_data %>% filter(num.artist == 1)
data_nart2 <- track_data %>% filter(num.artist == 2)
data_nart3 <- track_data %>% filter(num.artist == 3)

p_nart1 <- ggplot(data_nart1, aes(x=pop)) + geom_histogram()
p_nart2 <- ggplot(data_nart2, aes(x=pop)) + geom_histogram()
p_nart3 <- ggplot(data_nart3, aes(x=pop)) + geom_histogram()

p_nart1 
p_nart2
p_nart3
```

